Dark matter halo shapes in the Auriga simulations

### 2019-08-29 Document/paper.tex
last edits

% Abstract of the paper
\begin{abstract}
We present shape measurements of Milky Way sized dark matter halos at
redshift $z=0$ in a suite of 30 zoom simulations from the Auriga
project. 
We compare the results in full magnetohydrodynamics against dark
matter only simulations and find a strong influence of baryons 
in making dark matter haloes rounder at all radii compared to their dark 
matter only counterparts.
At distances $\lesssim 30$ kpc, rounder dark matter distributions
correlate with extended massive stellar discs and low core gas 
densities.  
We measure the alignment between the halo and the disc shapes at
different radii and find a high degree of alignment at all radii for most 
of the galaxies.
In some cases the alignment significantly changes as a function of
radius implying that the halo shape twists; 
this effect correlates with recently formed bulges and is almost
absent in the dark matter only simulations.  
In a comparison against observational constraints we find that $20\%$
of halos in our sample are consistent with observational results derived
from the Pal 5 stream that favours an almost spherical shape.
Including baryons is a required element to achieve this
level of agreement. In contrast, none of the simulations (neither dark matter
only nor with baryons) match the constraints derived from the
Sagittarius stream that favour an oblate dark matter halo.
\end{abstract}

@@

In this paper, we analyze the internal DM halo shape in thirty MW-type
galaxies  from the Auriga project, which are state-of-the-art hydrodynamical simulations of galaxy formation.
They have large enough numerical resolution, include an explicit
cosmological context, and have appropriate feedback physics to produce
realistic MW discs.
Furthermore, the relatively large number of simulated systems allows us
to obtain a handle on the statistical significance of our results.

@@

In particular, we measure the shape using the reduced inertia tensor
\citep{Allgood06},  
\begin{equation}
I_{ij} = \sum_k \frac{x_k^{(i)}x_k^{(j)}}{d^2_k},
\label{eq:inertia}
\end{equation}
where the particle positions are measured with respect to the minimum of the
gravitational potential in each halo and are weighted by the inverse of 
the k-th particle distance squared, $d_k^2 = \sum_{i=1}^3 x_k^{(i)} x_k^{(i)}$, giving each particle the same
weight independent of distance.

Diagonalization of this tensor yields the eigenvectors and
eigenvalues that approximate the dark matter halo with an ellipsoidal shape.
The axis lenghts of this ellipsoid, $a\geq b \geq c$, are the square
roots of the  eigenvalues, and the directions of the principal
axis are the corresponding eigenvectors.

@@

A common assumption in dynamical models of the MW dark matter halo is that
its minor axis is perfectly aligned with the angular momentym of the
stellar disc. 
Although this is a reasonable assumption due to the stability of
the galactic disc in this configuration in simplified models of isolated galaxies, this
might not perfectly hold in a full cosmological context. 
To examine the degree of validity of this assumption we study in this
subsection the alignment between the stellar disc's angular momentum
eigenvectors and the dark matter halo shape.

@@

In six halos the alignment changes significantly as a
function of radius produced by the twisting in the shape ellipsoids. 
In contrast, the DMO simulations show coherent halo shapes as a
function of radius without twisting. 
Because we are dealing with oblate halos $b/a\approx 1$, note that the
directions corresponding to the medium and major axis could be subject
to noise. 
For this reason, and in order to better quantify the twisting for each
halo, we take the standard deviation of the angle with the minor
axis, which should be less susceptible to noise.
With this measurement, twisting becomes a global property of the halo,
as the standard deviation is measured over all the different radii at
which we have quantified the alignment.

@@

The second part of our findings correspond to the disc-halo alignment.
The dominant trend is that of a strong alignment between
the disc and the dark matter halo shape at all radii. 
In six halos of our sample we find a chaning alignment as a function
of radius.
This implies a twisting in the halo shape as a function of
radius.
Considering that this effect is not present in our DMO simulations, it is
very likely that the detailed radial dependence of the alignments
between the disc and the halo depends on the hydrodynamics and
feedback implementations.

Twisting has been also reported as an uncommon feature in some 
simulated systems found in the literature. 
\cite{JingSuto02} reported it in their dark
matter only simulations as measured in the direction changes of large
and medium shape axis. 
They only had three high resolution MW-like halos and could not make a
statistical statement about the significance of this effect.
However, they see the twisting as an artifact resulting from high
values  of the $b/a$ ratios that make the determination of the medium
axis noisy, an effect that we have demonstrated can be discarded for
our simulations.
\cite{Bailin05} used seven hydrodynamic simulations to study the disc
halo alignment as a function of radius.
For radii smaller than $0.1R_{200}$ they find a strong global
alignment between the halo and the disc.
However, in some halos, at larger radii the angle between the
components twists resulting in a weaker correlation between the disk
and the halo shape. 
A more recent study by \cite{2015MNRAS.453..721V} used the EAGLE simulations to
report the median alignment between the stellar and dark matter
components in $1008$ galaxies around the MW mass range. 
Our results agree on the high degree of alignment on average as a
function of radius.
However, the authors do not mention whether they find any outliers
with the kind of twisting we report here.

@@

The results by \cite{Bovy16} are based on the same general approach but use
instead the GD-1 \citep{2006ApJ...641L..37G} and Pal 5 \citep{2009AJ....137.3378O}
streams to constraint the shape of the dark matter component of the
galacic halo potential. They use general models with many degrees of freedom for the galactic
potential in order to measure to what extent these two streams are sensitive
to the triaxiality of the dark matter halo component.
The DM component is written directly as a triaxial density profile
with coaxial ellipsoids, and the corresponding potential is found by
numerical integration. They find that the width of the Pal~5 stream gives $b/a\approx
1$, and therefore fix it to be $b/a=1$ exactly. Using this value they report their 
most stringent constraint of $c/a=0.93\pm0.16$ at a radius of $\approx 19$ kpc from the galactic
center.

@@

In this paper we have measured the shape of 30 isolated Milky Way sized
dark matter haloes simulated in the Auriga project using the zoom-in
technique. The halos were simulated using two different setups:
dark matter only (DMO) simulation, and full magnetohydrodynamics (MHD)
including star formation and feedback.
We applied the shape measurement algorithm by \cite{Allgood06} to the
dark matter halos of these simulations to quantify the halo shape as a
function of radius, and the degree of alignment between the angular
momentum in stellar disc and the dark matter halo shape.

@@

Using dark matter only simulations, \citet{VeraCiro11} suggested that
the current dark matter halo shape strongly correlates with the time
evolution of the halo as traced by the shape measured at the virial radius. 
This effect might largely be washed out and no longer hold once baryons are included. 
The opposite trend of triaxiality in MHD simulation as a function of
radius compared to the DMO simulations, together with the twisting effect in some of 
the halos, seems to suggest that memory of the historical buildup is at most poorly 
reflected in the shapes at $z=0$.

A more complete understanding of the influence of baryons on the
different properties we have measured will require at least two more elements.
First, a study on how the halo and the disc co-evolved as a
function of time. Second, a calibration of the effect of the cosmic web
(both dark matter and gaseous) on any evolutionary trend
\citep{2014MNRAS.443.1090F, 2017MNRAS.469..594B, 2019MNRAS.487.1607G}.
We also think  that the twisting density shells we find in
some of the halos are a feature that deserves further study, especially
in light of attempts to  constrain halo shape parameters with tidal stream data.  The
inclusion of a parameterization describing this degree of twisting
might relax the present conflicts between the observational inferences and the
numerical results.

>>>>

% Abstract of the paper
\begin{abstract}
We present shape measurements of Milky Way-sized dark matter halos at
redshift $z=0$ in a suite of 30 zoom simulations from the Auriga
project. 
We compare the results in full magnetohydrodynamics against dark
matter only simulations and find a strong influence of baryons 
in making dark matter haloes rounder at all radii compared to their dark 
matter only counterparts.
At distances $\lesssim 30$ kpc, rounder dark matter distributions
correlate with extended massive stellar discs and low core gas 
densities.  
We measure the alignment between the halo and the disc shapes at
different radii and find a high degree of alignment at all radii for most 
of the galaxies.
In some cases the alignment significantly changes as a function of
radius implying that the halo shape twists; 
this effect correlates with recently formed bulges and is almost
absent in the dark matter only simulations.  
In a comparison against observational constraints we find that $20\%$
of halos in our sample are consistent with observational results derived
from the Pal 5 stream that favours an almost spherical shape.
Including baryons is a required element to achieve this
level of agreement. In contrast, none of the simulations (neither dark matter
only nor with baryons) match the constraints derived from the
Sagittarius stream that favour an oblate dark matter halo.
\end{abstract}

@@

In this paper, we analyze the internal DM halo shape in thirty MW-type
galaxies  from the Auriga project, which are state-of-the-art hydrodynamical simulations of galaxy formation.
They have large enough numerical resolution, include an explicit
cosmological context, and have appropriate feedback physics to produce
realistic MW discs \citep{auriga,GBG18,GHF18}.
Furthermore, the relatively large number of simulated systems allows us
to obtain a handle on the statistical significance of our results.

@@

In particular, we measure the shape using the reduced inertia tensor
\citep{Allgood06},  
\begin{equation}
I_{ij} = \sum_k \frac{x_k^{(i)}x_k^{(j)}}{d^2_k},
\label{eq:inertia}
\end{equation}
where the particle positions are measured with respect to the minimum of the
gravitational potential in each halo and are weighted by the inverse of 
the $k$-th particle distance squared, $d_k^2 = \sum_{i=1}^3 x_k^{(i)} x_k^{(i)}$, giving each particle the same
weight independent of distance.

Diagonalization of this tensor yields the eigenvectors and
eigenvalues that approximate the dark matter halo with an ellipsoidal shape.
The axis lenghts of this ellipsoid, $a\geq b \geq c$, are the square
roots of the  eigenvalues, and the directions of the principal
axes are the corresponding eigenvectors.

@@

A common assumption in dynamical models of the MW dark matter halo is that
its minor axis is perfectly aligned with the angular momentum of the
stellar disc. 
Although this is a reasonable assumption due to the stability of
the galactic disc in this configuration in simplified models of isolated galaxies, this
might not perfectly hold in a full cosmological context. 
To examine the degree of validity of this assumption we study in this
subsection the alignment between the stellar disc's angular momentum
eigenvectors and the dark matter halo shape.

@@

In six halos the alignment changes significantly as a
function of radius produced by the twisting in the shape ellipsoids. 
In contrast, the DMO simulations show coherent halo shapes as a
function of radius without twisting with the exception of a single halo. 
Because we are dealing with oblate halos $b/a\approx 1$, note that the
directions corresponding to the medium and major axis could be subject
to noise. 
For this reason, and in order to better quantify the twisting for each
halo, we take the standard deviation of the angle with the minor
axis, which should be less susceptible to noise.
With this measurement, twisting becomes a global property of the halo,
as the standard deviation is measured over all the different radii at
which we have quantified the alignment.

@@

The second part of our findings concerns the disc-halo alignment.
The dominant trend is that of a strong alignment between
the disc and the dark matter halo shape at all radii. 
In six halos of our sample we find a changing alignment as a function
of radius.
This implies a twisting in the halo shape as a function of
radius.
Considering that this effect is virtually absent in our DMO simulations, it is
very likely that the detailed radial evolution of the alignments
between the disc and the halo depends on the hydrodynamics and
feedback implementations.

Twisting has been also reported as an uncommon feature in some 
simulated systems found in the literature. 
\cite{JingSuto02} reported it in their dark
matter only simulations as measured through direction changes of large
and medium shape axis. 
They only had three high resolution MW-like halos and could not make a
statistical statement about the significance of this effect.
However, they saw the twisting as an artifact resulting from high
values  of the $b/a$ ratios that make the determination of the medium
axis noisy, an effect that we have demonstrated can be discarded for
our simulations.
\cite{Bailin05} used seven hydrodynamic simulations to study the disc
halo alignment as a function of radius.
For radii smaller than $0.1R_{200}$ they find a strong global
alignment between the halo and the disc.
However, in some halos, at larger radii the angle between the
components twists resulting in a weaker correlation between the disk
and the halo shape. 
A more recent study by \cite{2015MNRAS.453..721V} used the EAGLE simulations to
report the median alignment between the stellar and dark matter
components in $1008$ galaxies around the MW mass range. 
Our results agree on the high degree of alignment on average as a
function of radius.
However, the authors do not mention whether they find any outliers
with the kind of twisting we report here.

@@

The results by \cite{Bovy16} are based on the same general approach but use
instead the GD-1 \citep{2006ApJ...641L..37G} and Pal 5 \citep{2009AJ....137.3378O}
streams to constraint the shape of the dark matter component of the
galacic halo potential. They use general models with many degrees of freedom for the galactic
potential in order to measure to what extent these two streams are sensitive
to the triaxiality of the dark matter halo component.
The DM component is written directly as a triaxial density profile
with coaxial ellipsoids, and the corresponding potential is found by
numerical integration. They find that the width of the Pal~5 stream gives $b/a\approx
1$, and therefore fix it to be $b/a=1$ exactly. Using this value they report their 
most stringent constraint of $c/a=0.93\pm0.16$ at a radius of $\approx 19$ kpc from the Galactic
center.

@@

In this paper we have measured the shape of 30 isolated Milky Way sized
dark matter haloes simulated in the Auriga project using the zoom-in
technique. The halos were simulated using two different setups:
dark matter only (DMO) simulation, and full magnetohydrodynamics (MHD)
including star formation and feedback.
We applied the shape measurement algorithm by \cite{Allgood06} to the
dark matter halos of these simulations to quantify the halo shape as a
function of radius, and the degree of alignment between the angular
momentum in the stellar disc and the dark matter halo shape.

@@

A more complete understanding of the influence of baryons on the
different properties we have measured will require at least two more elements.
First, a study on how the halo and the disc co-evolved as a
function of time. 
For instance, using dark matter only simulations, \citet{VeraCiro11} suggested that
the current dark matter halo shape strongly correlates with the time
evolution of the halo as traced by the shape measured at the virial radius. 
This effect might largely be washed out and no longer hold once baryons are included. 
The opposite trend of triaxiality in MHD simulation as a function of
radius compared to the DMO simulations, together with the twisting effect in some of 
the halos, seems to suggest that memory of the historical buildup is at most poorly 
reflected in the shapes at $z=0$.
The second element to take into account is the effect of the cosmic web
(both dark matter and gaseous) on any evolutionary trend
\citep{2014MNRAS.443.1090F, 2017MNRAS.469..594B, 2019MNRAS.487.1607G}.

We also think  that the twisting density shells we find in
some of the halos are a feature that deserves further study, especially
in light of attempts to  constrain halo shape parameters with tidal stream data.  The
inclusion of a parameterization describing this degree of twisting
might relax the present conflicts between the observational inferences and the
numerical results.
